---
title: "Mental Health Related Emergency Department Presentations (Australia)"
author: "Thomas Elton"
date: "Created on 27 Sep 2023; Updated on 19 Nov 2023"
output:
  html_document:
    fig_caption: yes
    number_sections: no
    self_contained: yes
    theme: flatly
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
---

> This markdown file may become outdated. For the most recent version, please see the GitHub repository here: https://github.com/tjelton/Matilda-Viz-2023-Dashboard-Entry

```{r message=FALSE, warning=FALSE}
# Load in libraries
library(tidyverse)
library(readxl)
library(stringr)
```

This document details the initial data analysis (IDA) of mental health emergency department data.

This is part of my analysis for the Matilda Viz 2023 competition.

# Data Source

The original data was sourced from Australian Government: Australian Institute of Health and Welfare. The original data can be downloaded [here](https://www.aihw.gov.au/mental-health/topic-areas/emergency-departments) and was last retrieved on the 28th of September 2023.

The data we will be looking at is on the sheet titled "Table ED.4". The name of the table is "Mental health-related emergency department presentations in public hospitals, by states and territories, 2014–15 to 2021–22."

***Citation:*** Australian Government: Australian Institute of Health and Welfare (2023). *Data tables: Mental health services provided in emergency departments 2021–22.* https://www.aihw.gov.au/mental-health/topic-areas/emergency-departments 

# Data Cleaning

The original data is in a wide format, which proves very difficult for plotting and analysis in R. Hence, the following code cleans the data into a long format. 

## Cleaning Process

```{r}
hospital_data <- read_excel("data/Mental-health-services-provided-in-emergency-departments-2122.xlsx", sheet = "Table ED.4", skip = 4)
```

```{r}
# Remove the blank rows that exist in the sheet for formatting.
# Remove the "Average annual change" column (this is column)
hospital_data_cleaned <- hospital_data %>% 
  filter(!is.na(Count)) %>%
  select(-13)
```

```{r}
# The following code converts the data into a long format (is is currently a wide format).
hospital_data_cleaned_long = hospital_data_cleaned %>%
  pivot_longer(
    cols = c(`2014–15`:`2021–22`),
    names_to = "year",
    values_to = "value"
  )
```

```{r}
# Rename year column so that we only focus on the first part of the year.
# e.g. For the year "2020-21", we only want to focus on "2020" (i.e. extract the first 4 characters)
for (value in 1:length(hospital_data_cleaned_long$year)) {
  hospital_data_cleaned_long$year[value] = stringr::str_extract(hospital_data_cleaned_long$year[value], "^.{4}")
}
```

```{r}
# Rename quarters to Q1 - Q4 for easier observing.
hospital_data_cleaned_long = hospital_data_cleaned_long %>%
  mutate(
    Quarter = case_when(
      grepl("January", Quarter) ~ "1",
      grepl("April", Quarter) ~ "2",
      grepl("July", Quarter) ~ "3",
      grepl("October", Quarter) ~ "4",
      grepl("All", Quarter) ~ "Year",
      TRUE ~ Quarter
    )
  )

# Remove the year rows as we can simply re-create that when needed.
hospital_data_cleaned_long = hospital_data_cleaned_long %>%
  filter(Quarter != "Year")
```

```{r}
# Create a column which will contain the year and quarter as a numerical value and a string.
hospital_data_cleaned_long = hospital_data_cleaned_long %>%
  mutate(
    # Default values.
    Time_Number = 0,
    Time_String = ""
  )

for (i in 1:length(hospital_data_cleaned_long$year)) {
  
  quarter = hospital_data_cleaned_long$Quarter[i]
  year = hospital_data_cleaned_long$year[i]
  
  # Set Time_String for the current row.
  string_1 = paste("Q", quarter, sep = "")
  string_2 = paste(string_1, year, sep = " ")
  hospital_data_cleaned_long$Time_String[i] = string_2
  
  # Set Time_Number for the current row.
  decimal = 0
  if (quarter == "2") {
    decimal = 0.25
  } else if (quarter == "3") {
    decimal = 0.5
  } else if (quarter == "4") {
    decimal = 0.75
  }
  hospital_data_cleaned_long$Time_Number[i] = as.integer(year) + decimal
}
```

```{r}
# Rename State and Territories to be acronyms.
hospital_data_cleaned_long = hospital_data_cleaned_long %>%
  
  # There must be some weird characters within the name of the states (realised when trying to merge later with a geojson file).
  # Solution is to rename states to a string where the characters are consistent.
  mutate(
    `State/Territory` = case_when(
      grepl("National", `State/Territory`) ~ "Australia",
      grepl("New", `State/Territory`) ~ "New South Wales",
      grepl("Western", `State/Territory`) ~ "Western Australia",
      grepl("Queensland", `State/Territory`) ~ "Queensland",
      grepl("Victoria", `State/Territory`) ~ "Victoria",
      grepl("Northern", `State/Territory`) ~ "Northern Territory",
      grepl("Capital", `State/Territory`) ~ "Australian Capital Territory",
      grepl("Tasmania", `State/Territory`) ~ "Tasmania",
      grepl("South", `State/Territory`) ~ "South Australia",
      TRUE ~ `State/Territory`
    )
  )
```

```{r}
# Rename columns.
hospital_data_cleaned_long = hospital_data_cleaned_long %>%
  rename(
    "Location" = "State/Territory",
    "Type" = "Type of presentation",
    "Year" = "year",
    "Value" =  "value"
    ) 
```

## The Final Cleaned Dataset

```{r}
write.csv(hospital_data_cleaned_long, "data/cleaned_mental_health_emergency_department.csv", row.names = FALSE)
str(hospital_data_cleaned_long)
```

# Initial Data Analysis

```{r}
hospital_data = read.csv("data/cleaned_mental_health_emergency_department.csv")
```

## Hospitalisations by Location

In order to plot the Australian states and territories, data was needed of the boundaries of each state and territory for plotting. These were found from here:

Hogan, R. (2014). *australian-states/states.geojson.* https://github.com/rowanhogan/australian-states/blob/master/states.geojson

*Note: To re-create this plot, you must donwload the states data from the GitHub link above!* 

```{r message=FALSE, warning=FALSE}
library(sf)
library(leaflet)
library(dplyr)

# Filter data to only focus on data of interest.
subset = hospital_data %>%
  filter(Location != "Australia") %>%
  filter(Time_String == "Q3 2021", Count == "Number", grepl("Mental", Type))
 
# Get spatial data.
# Spatial data downloaded from: https://github.com/rowanhogan/australian-states/blob/master/states.geojson
australia_sf <- sf::st_read("data/Australia_states.geojson")

# Join data.
merged = australia_sf %>% 
  dplyr::left_join(subset, by = c("STATE_NAME" = "Location"))
```

```{r}
# Credit: https://rstudio.github.io/leaflet/choropleths.html
map <- leaflet(merged)

# Make the colour palette for the leaflet be based upon the Value column.
pal <- colorQuantile("YlOrRd", domain = merged$Value)

# Labels for each state.
labels <- sprintf(
  "<strong>%s</strong><br/>%g Mental Health-Related Presentations",
  merged$STATE_NAME, merged$Value
) %>% lapply(htmltools::HTML)

# Create map.
map %>% addPolygons(
  fillColor = ~pal(Value),
  weight = 2,
  opacity = 1,
  color = "white",
  dashArray = "3",
  fillOpacity = 0.7,
  highlightOptions = highlightOptions(
    weight = 2,
    color = "#56575c",
    dashArray = "",
    fillOpacity = 0.7,
    bringToFront = TRUE),
  label = labels,
  labelOptions = labelOptions(
    style = list("font-weight" = "normal", padding = "3px 8px"),
    textsize = "15px",
    direction = "auto"))
```

## Trend in Hopsitalisations by Year

```{r}
# Get the type here because of weird string formatting.
# This is for "Mental health-related presentations" which contains characters that are atypical.
type = hospital_data$Type[1]

scatter_data = hospital_data %>% 
  filter(Count == "Number", Type == type) %>%
  filter(Year <= 2021) %>%
  group_by(Year, Location) %>%
  summarise(Value = sum(Value))
scatter_data$Year = as.numeric(scatter_data$Year)

fig = scatter_data %>% filter(Location != "Australia") %>%
  ggplot() +
  aes(x = Year, y = Value, colour = Location) + 
  geom_line() +
  geom_point() +
  expand_limits(y = 0) +
  labs(y = "Hospitalisations") +
  ggtitle("Australian Mental Health-Related Hospitalisations") +
  
  # Remove scientific notation.
  scale_y_continuous(labels = scales::comma) 

library(plotly)
ggplotly(fig)
```





