---
title: "Australian Deaths by Age and Sex (2019-2021)"
author: "Thomas Elton"
date: "Created on 28 Sep 2023; Updated on 19 Nov 2023"
output:
  html_document:
    fig_caption: yes
    number_sections: no
    self_contained: yes
    theme: flatly
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
---

> This markdown file may become outdated. For the most recent version, please see the GitHub repository here: https://github.com/tjelton/Matilda-Viz-2023-Dashboard-Entry

```{r message=FALSE, warning=FALSE}
# Load in libraries
library(tidyverse)
library(readxl)
library(stringr)
```

This document details the data cleaning and initial data analysis for the leading underlying causes of death by sex and age group for Australians from 2019 to 2021.

This is part of my analysis for the Matilda Viz 2023 competition.

# Data Source

The original data was sourced from Australian Government: Australian Institute of Health and Welfare. The original data can be downloaded [here](https://www.aihw.gov.au/reports/life-expectancy-deaths/deaths-in-australia/data) and was last retrieved on the 28th of September 2023.

In this analysis, we will be looking at the spreadsheet sheet titled "Table S3.2".

The original data source is not included on GitHub, and must be downloaded by the user from the following source.

# Data Cleaning

This data set is in a very poor format for plotting in R. One key challenge here is that there are multiple vertical sections in the spreadsheet for the different sexes: Males, Females, and Persons (all people combined).

## Cleaning Process

The way that cleaning this spreadsheet has been approached is by splitting the original data set into three; one for male, female, and all. Then we can merge these together.

```{r message=FALSE, warning=FALSE}
deaths_data = read_excel("data/AIHW-PHE-229-report-supplementary-tables.xlsx", sheet = "Table S3.2", skip = 3)
```

```{r}
# Remove columns which have no values (these separate the male, female and persons) columns in the data set.
deaths_data_cleaned = deaths_data %>%
  select(-c(`...6`, `...12`)) %>%
  
  # Remove rows at the bottom of the data set which are meta-data (this is the last 9 rows).
  head(-9)
```


```{r}
# Subset the male data.
male_df = deaths_data_cleaned %>%
  select(`Age group and rank...1`:`Age-specific rate (per 100,000)...5`)

# Rename columns
male_df <- male_df %>%
  rename(
    "Age_Group" = "Age group and rank...1",
    "Cause_of_Death" = "Cause of death...2",
    "Count" = "Number...3",
    "Percentage_of_Deaths" = "Per cent...4",
    "Rate_Per_100000_Age_Specific" = "Age-specific rate (per 100,000)...5"
  )
```

```{r}
# Cleaning function
clean_sex_subset <- function(df, sex) {
  
  # We want to keep the "All causes" row in each data set.
  # Hence, in the "Age_Group" column we will give it a rank of 21 for now.
  # This is because we are about to delete it and replace each ranking with their Age Group.
  for (i in 1:length(df$Age_Group)) {
    
    # Skip over NA values.
    if (is.na(df$Cause_of_Death[i])) {
      next
    }
    if ("All causes" == df$Cause_of_Death[i]) {
      df$Age_Group[i] = "21"
      
      # The rows for all causes should have the percent being 100.
      df$Percentage_of_Deaths[i] = "100"
    }
  }
  
  # Remove all rows in Age_Group that are NA
  df = df %>%
    filter(!is.na(Age_Group))
  
  # Remove rankings from the Age_Group column.
  # The age groups from the first row of each section should carry through.
  age_group = "0-1"
  # Skip first row which just contains the string "Under 1"
  for (i in 2:length(df$Age_Group)) {
    
    current_value = df$Age_Group[i]
    
    # Update the age_group variable is a "-" character is seen.
    if (grepl("–", current_value) || grepl("\\+", current_value) || grepl("All", current_value)) {
      age_group = current_value
      next
    }
    
    df$Age_Group[i] = age_group
  }
  
  # Remove all rows where cause of death is 0.
  df = df %>%
    filter(!is.na(Cause_of_Death))
  
  # Add a Gender column where the gender is specified as a function parameter
  df = df %>% 
    mutate(Sex = sex)
  
  
  # Remove the content in brackets in the Cause_of_Death column.
  df = df %>%
    mutate(Cause_of_Death = gsub("\\(.*\\)", "", Cause_of_Death))
  
  return(df)
}
```

```{r}
# Clean male data subset.
male_df = clean_sex_subset(male_df, "Male")
```

```{r}
# Clean female data subset.
female_df = deaths_data_cleaned %>%
  select(`Age group and rank...7`:`Age-specific rate (per 100,000)...11`)

# Rename columns
female_df <- female_df %>%
  rename(
    "Age_Group" = "Age group and rank...7",
    "Cause_of_Death" = "Cause of death...8",
    "Count" = "Number...9",
    "Percentage_of_Deaths" = "Per cent...10",
    "Rate_Per_100000_Age_Specific" = "Age-specific rate (per 100,000)...11"
  )

female_df = clean_sex_subset(female_df, "Female")
```

```{r}
# Clean all data subset.
all_df = deaths_data_cleaned %>%
  select(`Age group and rank...13`:`Age-specific rate (per 100,000)...17`)

# Rename columns
all_df <- all_df %>%
  rename(
    "Age_Group" = "Age group and rank...13",
    "Cause_of_Death" = "Cause of death...14",
    "Count" = "Number...15",
    "Percentage_of_Deaths" = "Per cent...16",
    "Rate_Per_100000_Age_Specific" = "Age-specific rate (per 100,000)...17"
  )

all_df = clean_sex_subset(all_df, "All")
```

```{r}
# Combine the different subsets back into one data set.
deaths_cleaned = rbind(male_df, female_df, all_df)
```

## The Final Cleaned Dataset

```{r}
write.csv(deaths_cleaned, "data/death_data_cleaned.csv")
str(deaths_cleaned)
```

# Initial Data Analaysis

```{r}
library(DT)
library(formattable)
data = read.csv("data/death_data_cleaned.csv") %>% 
  select(-X) %>%
  filter(Cause_of_Death != "All causes") %>%
  rename(
    "Age Group" = "Age_Group",
    "Cause of Death" = "Cause_of_Death",
    "Deaths" = "Count",
    "Percentage of Deaths" = "Percentage_of_Deaths",
    "Rate per 100k" = "Rate_Per_100000_Age_Specific"
  ) %>%
  
  # Add a rank column to rank causes within an age group and sex
  group_by(`Age Group`, Sex) %>%
  mutate(Rank = dense_rank(desc(Deaths)))
```

```{r}
# Table showing different causes of death, and highlighting the row that says suicide.
subset = data %>%
  filter(grepl("15", `Age Group`), grepl("Male", Sex), `Cause of Death` != "All causes") %>%
  
  # Make a ranking column (rank based off number of deaths).
  mutate(Rank = dense_rank(desc(Deaths))) %>%
  
  # Re-order columns to make Rank, Deaths, Percentage of Deaths and Rate per 100k the first columns.
  select(Rank, Deaths,`Percentage of Deaths`, `Rate per 100k`, `Cause of Death`, `Age Group`, Sex)

datatable(subset, rownames = FALSE) %>% formatStyle(
  'Cause of Death',
  target = 'row',
  backgroundColor = styleEqual(c("Suicide "), c('#fabdb9'))
)
```

```{r message=FALSE, warning=FALSE}
# Filter so that only the death types closest to suicide are plot (avoids too much being plotted)
subset = data %>%
  filter(grepl("15", `Age Group`), grepl("Male", Sex), `Cause of Death` != "All causes")

subset_2 = subset %>%
  filter(`Cause of Death` == "Suicide ")

# Get 2 ranks ahead of suicide and 2 ranks below suicide
suicide_rank = subset %>%
  filter(`Cause of Death` == "Suicide ") %>%
  select(Rank) %>%
  pull()
bottom_bound = suicide_rank - 2
upper_bound = suicide_rank + 2

# Filter to the ranks of interest.
subset_for_plotting = subset %>%
  filter(Rank >= bottom_bound & Rank <= upper_bound) %>%
  mutate(`Cause of Death` = paste(str_trunc(`Cause of Death`, 25), "(Rank ", as.character(Rank), ")", sep = "")) %>%
  
  # Add colours for custom colouring (every bar will be grey, except for suicide which will be red).
  mutate(
    colour = case_when(
      grepl("Suicide", `Cause of Death`) ~ "rgba(250,189,185,1)",
      TRUE ~ "rgba(204,204,204,1)"
    )
  )

# Create column graph
library(plotly)
fig <- plot_ly(
  x = subset_for_plotting$`Cause of Death`,
  y = subset_for_plotting$Deaths,
  type = "bar",
  marker = list(color = subset_for_plotting$colour)
) %>% 
  layout(yaxis = list(title = "Deaths"), xaxis = list(title = "Cause of Death"))

fig
```


